---
title: "An Introduction to Restricted Mean Survival Time (RMST)"
author: "Akif Mustafa"
date: 2026-06-01
categories: [Survival analysis, Clinical trials, R]
description: "An introduction to RMST, why it matters when proportional hazards fail, and how to estimate it in R."
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: false
---

```{r setup}
#| include: false
library(survival)
library(survminer)
library(ggplot2)
library(survRM2)
library(dplyr)
library(patchwork)
library(knitr)
library(kableExtra)
```

## Introduction

The **Cox proportional hazards (CPH) model** is the go-to model for survival analysis. It is popular for good reasons:

-   It is semi-parametric --- no distributional assumption needed.
-   It handles multiple predictor variables.
-   It gives an easy-to-interpret summary: the **hazard ratio (HR)**.
-   It handles censoring naturally.

But it makes one strong assumption: **proportional hazards (PH)**.

This means the hazard of the event in two groups must always remain proportional throughout the study period. Everything works well when this holds.

**What if it does not hold?**

When hazards are not proportional --- for example, when survival curves cross, when there is a delayed treatment effect, or when the benefit diminishes over time --- a single HR becomes misleading. It no longer summarizes the treatment effect well.

```{r non-ph-illustration}
#| fig-cap: "Potential hazard scenarious"
#| fig-width: 10
#| fig-height: 8
#| echo: false

t <- seq(0, 50, length.out = 500)
c1 <- exp(-0.05 * t)       # Control: Constant hazard of 0.05
t1 <- exp(-0.025 * t)      # Treatment: Constant hazard of 0.025
c2 <- exp(-0.03 * t)       # Control: Constant hazard
t2 <- exp(-0.002 * (t^2))  # Treatment: Weibull survival (hazard increases)
c3 <- exp(-0.05 * t)
t3 <- ifelse(t <= 15, 
             exp(-0.05 * t), 
             exp(-0.05 * 15) * exp(-0.015 * (t - 15)))
c4 <- exp(-0.05 * t)
t4 <- ifelse(t <= 15, 
             exp(-0.015 * t), 
             exp(-0.015 * 15) * exp(-0.1 * (t - 15)))
df <- data.frame(
  Time = rep(t, 8),
  Survival = c(c1, t1, c2, t2, c3, t3, c4, t4),
  Group = rep(rep(c("Control", "Treatment"), each = length(t)), 4),
  Scenario = rep(c("PH Holding", 
                   "Crossing Curves", 
                   "Delayed Treatment Effect", 
                   "Diminishing Benefit"), 
                 each = 2 * length(t))
)
df$Scenario <- factor(df$Scenario, levels = c(
  "PH Holding", 
  "Crossing Curves", 
  "Delayed Treatment Effect", 
  "Diminishing Benefit"
))
p <- ggplot(df, aes(x = Time, y = Survival, color = Group, linetype = Group)) +
  geom_line(linewidth = 1) +
  # Create the 2x2 grid based on the 'Scenario' column
  facet_wrap(~ Scenario, nrow = 2) +
  scale_color_manual(values = c("Control" = "#E41A1C", "Treatment" = "#377EB8")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Time",
       y = "Survival Probability") +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "#f0f0f0")
  )
print(p)
```

In these situations, there are several analytical alternatives to the standard Cox model:

-   Piecewise Cox model
-   Cox model with time-varying coefficients
-   Royston-Parmar parametric model
-   Accelerated failure time (AFT) model
-   Cure models

**Restricted Mean Survival Time (RMST)** is one such alternative. It is practical, interpretable, and widely used when the proportional hazards assumption is violated.

------------------------------------------------------------------------

## What is RMST?

The **Restricted Mean Survival Time (RMST)** up to a time point $\tau$ is:

-   The **average time** a person survives from time 0 to time $\tau$.
-   Equivalently, the **area under the survival curve** $S(t)$ from 0 to $\tau$.

Mathematically:

$$
\text{RMST}(\tau) = E[\min(T, \tau)] = \int_0^{\tau} S(t)\, dt
$$

Two key points:

-   **"Restricted"** means we only look up to a pre-specified horizon $\tau$ (e.g., 5 years), not to infinity.
-   It is a **non-parametric** summary of the whole survival curve up to $\tau$. No proportional hazards assumption. No specific model required.

### Visualizing RMST: Area Under the Curve

```{r rmst-auc-plot}
#| fig-cap: "RMST is the area under the survival curve S(t) from time 0 to τ"
#| fig-width: 7
#| fig-height: 4.5

set.seed(1)
tau <- 48

t_seq <- seq(0, 60, by = 0.1)
s_t   <- exp(-0.03 * t_seq)

df_line <- data.frame(time = t_seq, surv = s_t)
df_shade <- data.frame(
  time = c(0, t_seq[t_seq <= tau], tau),
  surv = c(0, s_t[t_seq <= tau],  0)
)

ggplot() +
  geom_polygon(data = df_shade, aes(time, surv),
               fill = "steelblue", alpha = 0.35) +
  geom_line(data = df_line, aes(time, surv),
            color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = tau, linetype = "dashed", color = "black") +
  annotate("text", x = tau + 1.5, y = 0.85, label = expression(tau),
           size = 5, hjust = 0) +
  annotate("text", x = tau / 2.5, y = 0.22,
           label = "RMST = Area\nunder S(t)\nup to τ",
           size = 3.8, color = "steelblue4", fontface = "bold") +
  labs(
    title    = "Restricted Mean Survival Time (RMST)",
    subtitle = paste0("Shaded area = RMST up to τ = ", tau, " months"),
    x        = "Time (months)",
    y        = "Survival probability S(t)"
  ) +
  scale_x_continuous(breaks = seq(0, 60, 12)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  theme_classic(base_size = 12)
```

### RMST for Two Groups

For two groups (e.g., treatment vs. control), the **difference in RMST** is the area *between* the two survival curves from 0 to $\tau$.

It tells you: **on average, how much extra survival time does the treatment provide up to time** $\tau$?

```{r rmst-two-groups-plot}
#| fig-cap: "RMST difference = shaded area between the two survival curves up to τ"
#| fig-width: 7
#| fig-height: 4.5

set.seed(2)
tau <- 48

t_seq  <- seq(0, 60, by = 0.1)
s_trt  <- exp(-0.020 * t_seq)
s_ctrl <- exp(-0.035 * t_seq)

df_two <- data.frame(
  time    = t_seq,
  Treated = s_trt,
  Control = s_ctrl
) |> tidyr::pivot_longer(-time, names_to = "Group", values_to = "Survival")

df_between <- data.frame(
  time  = c(t_seq[t_seq <= tau], rev(t_seq[t_seq <= tau])),
  surv  = c(s_trt[t_seq <= tau],  rev(s_ctrl[t_seq <= tau]))
)

ggplot() +
  geom_polygon(data = df_between, aes(time, surv),
               fill = "gold", alpha = 0.5) +
  geom_line(data = df_two, aes(time, Survival, color = Group),
            linewidth = 1.2) +
  geom_vline(xintercept = tau, linetype = "dashed", color = "black") +
  annotate("text", x = tau + 1.5, y = 0.9, label = expression(tau),
           size = 5, hjust = 0) +
  annotate("text", x = 15, y = 0.55,
           label = "RMST\nDifference",
           size = 3.8, color = "darkgoldenrod", fontface = "bold") +
  scale_color_manual(values = c("Control" = "tomato", "Treated" = "steelblue")) +
  labs(
    title    = "RMST Difference Between Two Groups",
    subtitle = "Shaded area = difference in RMST up to τ",
    x        = "Time (months)",
    y        = "Survival probability S(t)"
  ) +
  scale_x_continuous(breaks = seq(0, 60, 12)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

## Why RMST? Limitations of the Cox Model

The Cox model gives hazard ratios. HRs are useful, but they have real limitations.

### PH Assumption

The Cox model requires **proportional hazards**. If hazards are not proportional (crossing curves, delayed effects, diminishing benefits), a single HR is misleading.

> In immuno-oncology, for example, survival curves often separate late or cross altogether. The PH assumption frequently fails here.

### HR is Relative, Not Absolute

HR = 0.6 means a 40% lower hazard. But what does that mean in **months of life gained**? That depends on the baseline hazard --- a detail HR does not convey.

Clinicians and patients often more interested in:

> *"How much extra time do I get, on average?"*

than about a hazard ratio.

### HR is Model-Based

The Cox HR is a model-based summary. If the model is wrong --- due to non-proportional hazards or misspecification --- the HR can be biased and difficult to interpret.

### How RMST Addresses These Limitations

| Issue                       | Cox HR      | RMST                   |
|-----------------------------|-------------|------------------------|
| Requires PH assumption      | ✅ Yes      | ❌ No                  |
| Absolute effect size        | ❌ No       | ✅ Yes (time gained)   |
| Model-free                  | ❌ No       | ✅ Yes (from KM curve) |
| Easy clinical communication | ⚠️ Moderate | ✅ High                |

> **RMST is not a replacement for the Cox model in all settings.** It is a complement and an alternative --- especially when PH fails or an absolute effect size matters more.

## When to Use RMST?

### Use RMST When:

**1. Non-proportional hazards (NPH) are present**

-   Crossing survival curves
-   Delayed treatment effect (early overlap, later separation)
-   Diminishing benefit (early benefit that later fades)

**2. An absolute effect is more meaningful than a relative one**

-   *"Patients in the active arm survive on average 3 months longer over 5 years."*
-   Health economics: quantifying life-years gained within a fixed horizon.

**3. You want a model-free summary**

-   No reliance on a correctly specified Cox model.
-   Simple, robust comparison of two survival curves up to $\tau$.

**4. As a supplement to HRs and log-rank tests**

-   Even when PH holds, RMST can be reported alongside the HR to provide an absolute effect estimate.
-   Many recent clinical trials in oncology report both HR and RMST difference.

### You Probably Do Not Need RMST When:

-   PH holds well, **and**
-   HR is sufficient for your scientific or clinical question, **and**
-   There is no strong demand for an absolute "time gained" interpretation.

## Estimation

### Single Group

The steps are straightforward:

1.  Estimate the survival curve $S(t)$ using the **Kaplan--Meier** estimator.
2.  Choose a clinically relevant truncation time $\tau$ (e.g., 5 years). Pre-specify this before the analysis.
3.  Compute the area under the KM curve from 0 to $\tau$ using the **trapezoidal rule** (sum of rectangles under the step function).
4.  Estimate the **standard error** using a Greenwood-type formula.
5.  Construct **confidence intervals** using asymptotic normality (for large samples).

### Two-Group Comparison

Let $S_1(t)$ and $S_0(t)$ be the KM curves in the treatment and control groups.

$$
\text{RMST}_1(\tau) = \int_0^{\tau} S_1(t)\, dt \qquad
\text{RMST}_0(\tau) = \int_0^{\tau} S_0(t)\, dt
$$

Three key summary measures:

**1. Difference in RMST (most common)**

$$
\Delta(\tau) = \text{RMST}_1(\tau) - \text{RMST}_0(\tau)
= \int_0^{\tau} [S_1(t) - S_0(t)]\, dt
$$

*Interpretation*: On average, how many extra months does treatment provide by time $\tau$?

**2. Ratio of RMST**

$$
\frac{\text{RMST}_1(\tau)}{\text{RMST}_0(\tau)}
$$

*Interpretation*: Relative increase in average survival time up to $\tau$.

**3. Ratio of Restricted Mean Time Lost (RMTL)**

$$
\text{RMTL}(\tau) = \tau - \text{RMST}(\tau)
$$

This is the average time **lost** due to the event before $\tau$.

$$
\text{RMTL ratio} = \frac{\text{RMTL}_1(\tau)}{\text{RMTL}_0(\tau)}
$$

*Interpretation*: Useful when the focus is on time lost rather than time lived.

**Inference** (difference and ratio):

-   Based on asymptotic normality with Greenwood-based variance estimates.
-   Provides standard errors, confidence intervals, and p-values.

------------------------------------------------------------------------

## Choosing $\tau$

The choice of $\tau$ matters. Some guidance:

-   $\tau$ should be **clinically chosen and pre-specified** before the analysis (e.g., 5-year RMST in a cancer trial).
-   It must fall **within the observed follow-up** where enough events have occurred and the KM estimate is reliable (i.e., do not extrapolate beyond the last event time).
-   A common practical rule: set $\tau$ at or before the **minimum of the two groups' last event times**, or at a **clinically meaningful landmark** such as 1, 2, or 5 years.
-   Some methods choose $\tau$ adaptively, but a **fixed, pre-specified** $\tau$ is preferred for transparency and interpretability.

```{r tau-choice-illustration}
#| fig-cap: "Effect of τ choice on RMST estimation. The shaded areas show RMST
#|   for three different values of τ."
#| fig-width: 9
#| fig-height: 3.5
#| echo: false

t_seq <- seq(0, 60, by = 0.1)
s_t   <- exp(-0.028 * t_seq)

tau_vals  <- c(24, 36, 48)
fill_cols <- c("#AED6F1", "#5DADE2", "#1A5276")

plots <- lapply(seq_along(tau_vals), function(i) {
  tau_i <- tau_vals[i]
  df_shade_i <- data.frame(
    time = c(0, t_seq[t_seq <= tau_i], tau_i),
    surv = c(0, s_t[t_seq <= tau_i],   0)
  )
  rmst_val <- sum(diff(c(0, t_seq[t_seq <= tau_i])) *
                    s_t[t_seq <= tau_i]) |> round(1)

  ggplot() +
    geom_polygon(data = df_shade_i, aes(time, surv),
                 fill = fill_cols[i], alpha = 0.6) +
    geom_line(data = data.frame(time = t_seq, surv = s_t),
              aes(time, surv), color = "steelblue", linewidth = 1) +
    geom_vline(xintercept = tau_i, linetype = "dashed", color = "black") +
    annotate("text", x = tau_i + 1, y = 0.9,
             label = paste0("τ = ", tau_i), size = 3.5, hjust = 0) +
    annotate("text", x = tau_i / 2.8, y = 0.15,
             label = paste0("RMST ≈ ", rmst_val),
             size = 3.2, color = "white", fontface = "bold") +
    labs(x = "Time", y = "S(t)",
         title = paste0("τ = ", tau_i, " months")) +
    scale_x_continuous(limits = c(0, 62), breaks = seq(0, 60, 20)) +
    scale_y_continuous(limits = c(0, 1)) +
    theme_classic(base_size = 10)
})

plots[[1]] + plots[[2]] + plots[[3]]
```

------------------------------------------------------------------------

## Example in R

Here we simulate a delayed treatment effect:

-   Before 12 months: treatment and control have similar hazard.
-   After 12 months: treatment hazard becomes lower.

This violates proportional hazards.

### Simulate Data

```{r}
#| label: simulate-nph-data

library(survival)
library(survRM2)
library(ggplot2)
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)

set.seed(999)

n <- 600
arm <- rbinom(n, 1, 0.5)  # 0 = control, 1 = treatment

# Control: constant hazard throughout
t_control <- rexp(n, rate = 0.05)

# Treatment: same hazard before 12 months, lower hazard after 12 months
t1 <- rexp(n, rate = 0.05)
t_treat <- ifelse(t1 < 12, t1, 12 + rexp(n, rate = 0.02))

time_event <- ifelse(arm == 0, t_control, t_treat)

# Administrative censoring at 36 months
time   <- pmin(time_event, 36)
status <- as.integer(time_event <= 36)

dat <- tibble(
  time   = time,
  status = status,
  arm    = factor(arm, levels = c(0, 1), labels = c("Control", "Treatment"))
)

# Summary table
dat |>
  group_by(arm) |>
  summarise(
    N            = n(),
    Events       = sum(status),
    Censored     = N - Events,
    Max_followup = round(max(time), 1),
    .groups      = "drop"
  ) |>
  kable(
    caption   = "Summary of simulated dataset",
    col.names = c("Group", "N", "Events", "Censored", "Max Follow-up (months)")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### Kaplan--Meier Curves

```{r}
#| label: km-plot
#| fig-cap: "Kaplan–Meier curves for simulated delayed treatment effect data"
#| fig-width: 7
#| fig-height: 4.5

fit_km <- survfit(Surv(time, status) ~ arm, data = dat)

# Extract KM step data for ggplot
km_df <- data.frame(
  time   = fit_km$time,
  surv   = fit_km$surv,
  strata = rep(names(fit_km$strata), fit_km$strata)
) |>
  mutate(arm = sub("arm=", "", strata))

# Add time = 0 rows for each group
time0 <- data.frame(
  time   = 0,
  surv   = 1,
  strata = NA,
  arm    = c("Control", "Treatment")
)

km_plot_df <- bind_rows(time0, km_df) |>
  arrange(arm, time)

ggplot(km_plot_df, aes(x = time, y = surv, color = arm)) +
  geom_step(linewidth = 1) +
  scale_color_manual(
    name   = "Group",
    values = c("Control" = "tomato", "Treatment" = "steelblue")
  ) +
  scale_x_continuous(limits = c(0, 36), breaks = seq(0, 36, 6)) +
  scale_y_continuous(limits = c(0, 1),  breaks = seq(0, 1, 0.2)) +
  labs(
    title    = "Kaplan–Meier Curves: Delayed Treatment Effect",
    subtitle = "Curves overlap early and diverge after ~12 months",
    x        = "Time (months)",
    y        = "Survival probability"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

The curves initially overlap, reflecting similar early hazards. After 12 months they diverge as the treatment arm benefits from a lower hazard. This pattern suggests a violation of the proportional hazards assumption.

### Check the Proportional Hazards Assumption

#### First, lets fit the cox PH model

```{r}
cox_fit <- coxph(Surv(time, status) ~ arm, data = dat)
summary(cox_fit)
```

#### Now, check the PH assumption using schoenfeld test

```{r}
#| label: schoenfeld-test

zph <- cox.zph(cox_fit)

# Print test result as a neat table
as.data.frame(zph$table) |>
  kable(
    caption = "Schoenfeld residuals test for proportional hazards",
    digits  = 4
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

> A p-value below 0.05 on the Schoenfeld test indicates a violation of the PH assumption.

#### Plot the Schoenfeld residuals for visual inspection

```{r}
#| label: schoenfeld-plot
#| fig-cap: "Scaled Schoenfeld residuals. A flat line suggests PH holds; a trend suggests violation."
#| fig-width: 7
#| fig-height: 4

plot(zph, main = "Schoenfeld Residuals Over Time")
abline(h = 0, lty = 2, col = "grey50")
```

> A non-linear trend in the residuals plot confirms this. PH is violated here --- making RMST a better choice.

### Compute RMST

We use the `survRM2` package and set $\tau = 36$ months.

```{r}
#| label: rmst-compute

tau_val <- 36

rm <- rmst2(
  time   = dat$time,
  status = dat$status,
  arm    = as.integer(dat$arm == "Treatment"),  # 1 = treatment, 0 = control
  tau    = tau_val
)

rm
```

#### Results Tables

```{r}
#| label: rmst-per-group-table

arm1_vec <- unlist(rm$RMST.arm1)
arm0_vec <- unlist(rm$RMST.arm0)

rmst_trt_est  <- as.numeric(arm1_vec[1])
rmst_trt_lo   <- as.numeric(arm1_vec[2])
rmst_trt_hi   <- as.numeric(arm1_vec[3])

rmst_ctrl_est <- as.numeric(arm0_vec[1])
rmst_ctrl_lo  <- as.numeric(arm0_vec[2])
rmst_ctrl_hi  <- as.numeric(arm0_vec[3])

rmst_summary <- data.frame(
  Group = c("Treatment", "Control"),
  RMST  = c(rmst_trt_est,  rmst_ctrl_est),
  Lower = c(rmst_trt_lo,   rmst_ctrl_lo),
  Upper = c(rmst_trt_hi,   rmst_ctrl_hi)
)

rmst_summary |>
  kable(
    caption   = paste0("RMST estimates up to τ = ", tau_val, " months"),
    col.names = c("Group", "RMST (months)", "95% CI Lower", "95% CI Upper"),
    digits    = 2
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

#### Contrast Table

```{r}
#| label: rmst-contrast-table

contrast_df <- as.data.frame(rm$unadjusted.result)
contrast_df$Measure <- c(
  "RMST Difference (Treatment - Control)",
  "RMST Ratio (Treatment / Control)",
  "RMTL Ratio (Treatment / Control)"
)
contrast_df <- contrast_df[, c("Measure", "Est.", "lower .95", "upper .95", "p")]

contrast_df |>
  kable(
    caption   = "RMST contrast measures",
    col.names = c("Measure", "Estimate", "95% CI Lower", "95% CI Upper", "P-value"),
    digits    = c(0, 2, 2, 2, 4)
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### Visualize RMST

```{r}
#| label: rmst-visual
#| fig-cap: "KM curves with shaded RMST areas up to τ = 36 months"
#| fig-width: 7
#| fig-height: 5

# Fit separate KM curves for each arm
km_trt  <- survfit(Surv(time, status) ~ 1, data = filter(dat, arm == "Treatment"))
km_ctrl <- survfit(Surv(time, status) ~ 1, data = filter(dat, arm == "Control"))

# Helper: convert survfit object to a step-function data frame
km_to_df <- function(km_obj, group_label) {
  data.frame(
    time  = c(0, km_obj$time),
    surv  = c(1, km_obj$surv),
    group = group_label
  )
}

df_trt  <- km_to_df(km_trt,  "Treatment")
df_ctrl <- km_to_df(km_ctrl, "Control")

# Helper: build a closed polygon for shading up to tau
make_shade <- function(df, tau_limit) {
  # Keep only time points up to tau
  df_trunc <- df[df$time <= tau_limit, ]
  # Add a closing point at tau with the last survival value
  df_trunc <- rbind(df_trunc,
                    data.frame(time  = tau_limit,
                               surv  = tail(df_trunc$surv, 1),
                               group = df_trunc$group[1]))
  # Close the polygon back to zero
  data.frame(
    time  = c(0,             df_trunc$time, tau_limit, 0),
    surv  = c(0,             df_trunc$surv, 0,         0),
    group = df_trunc$group[1]
  )
}

shade_trt  <- make_shade(df_trt,  tau_val)
shade_ctrl <- make_shade(df_ctrl, tau_val)

ggplot() +
  # Shaded RMST areas
  geom_polygon(data = shade_trt,
               aes(time, surv),
               fill = "steelblue", alpha = 0.25) +
  geom_polygon(data = shade_ctrl,
               aes(time, surv),
               fill = "tomato", alpha = 0.25) +
  # KM step curves
  geom_step(data = df_trt,
            aes(time, surv, color = "Treatment"),
            linewidth = 1) +
  geom_step(data = df_ctrl,
            aes(time, surv, color = "Control"),
            linewidth = 1) +
  # Tau line
  geom_vline(xintercept = tau_val, linetype = "dashed", color = "black") +
  annotate("text",
           x = tau_val - 1, y = 0.05,
           label = paste0("\u03c4 = ", tau_val, " months"),
           hjust = 1, size = 3.5) +
  scale_color_manual(
    name   = "Group",
    values = c("Treatment" = "cyan3", "Control" = "tomato")
  ) +
  scale_x_continuous(limits = c(0, 36), breaks = seq(0, 36, 6)) +
  scale_y_continuous(limits = c(0, 1),  breaks = seq(0, 1, 0.2)) +
  labs(
    title    = "RMST Visualization: KM Curves with Shaded Areas",
    subtitle = paste0("Shaded area = RMST up to \u03c4 = ", tau_val, " months"),
    x        = "Time (months)",
    y        = "Survival probability S(t)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

### Interpretation

```{r}
#| label: extract-results
#| echo: false

rmst_trt_est  <- round(rm[["RMST.arm1"]][["rmst"]][["Est."]])
rmst_ctrl_est <- round(rm[["RMST.arm0"]][["rmst"]][["Est."]])

row_id <- grep("RMST.*arm=1.*arm=0", rownames(rm$unadjusted.result))
rmst_diff    <- round(rm$unadjusted.result[1, "Est."], 1)
rmst_diff_lo <- round(rm$unadjusted.result[1, "lower .95"], 2)
rmst_diff_hi <- round(rm$unadjusted.result[1, "upper .95"], 3)
rmst_pval    <- round(rm$unadjusted.result[1, "p"], 4)
```

::: callout-note
## Key Results

-   **Treatment group**: RMST = `r rmst_trt_est` months up to $\tau$ = `r tau_val` months.
-   **Control group**: RMST = `r rmst_ctrl_est` months up to $\tau$ = `r tau_val` months.
-   **RMST Difference**: `r rmst_diff` months (95% CI: `r rmst_diff_lo` to `r rmst_diff_hi`; p = `r rmst_pval`).

**Plain language**: On average, patients in the treatment group survive `r rmst_diff` months longer within the first `r tau_val` months compared to the control group.

This is a clinically meaningful and model-free summary of the treatment benefit --- no proportional hazards assumption required.
:::

## References

-   Uno H, et al. (2014). Moving beyond the hazard ratio in quantifying the between-group difference in survival analysis. *Journal of Clinical Oncology*.
-   Royston P, Parmar MKB (2013). Restricted mean survival time: an alternative to the hazard ratio for the design and analysis of randomized trials with a time-to-event outcome. *BMC Medical Research Methodology*.
-   Trinquart L, et al. (2016). Comparison of treatment effects measured by the hazard ratio and by the ratio of restricted mean survival times in oncology randomized controlled trials. *Journal of Clinical Oncology*.
